Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-05-02 ms_pre_post.Rmd
Edit through Methods

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply 'inbreeding strategies' because the model did not preclude evolution of inbreeding preference), in a small focal population of females and males by tracking the dynamics of alleles underlying reproductive strategies expressed by females. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and allowing for effects of mutation, gene flow, drift, and selection. All individuals have 10 physically unlinked diploid loci, and therefore 20 alleles, underlying each of three reproductive strategy traits (i.e., 30 diploid loci in total): tendency for polyandry ($P_{a}$, i.e., 'polyandry' alleles), pre-copulatory inbreeding strategy ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, i.e., 'fertilisation' alleles). All alleles can take the value of any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to determine genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy, respectively, whereby an individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. Phenotype values $M_{p} = M_{g}$ and $F_{p} = F_{g}$, where negative and positive values cause inbreeding avoidance and preference, respectively (see details of mating and fertilisation strategies below). In contrast, $P_{p} = P_{g}$ only if $P_{g} \geq 0$ to cause increasingly greater tendency for polyandry, while $P_{p} = 0$ if $P_{g} < 0$ because females cannot choose a negative number of mates (see details of polyandry below). Polyandry is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998; @Duthie] that is influenced by continuous genetic variation but only expressed when $P_{g} > 0$. Any negative $P_{g}$ value therefore generates phenotypic monandry, while positive $P_{g}$ values can lead to different degrees of realised polyandry (i.e., different numbers of mates, $N_{males}$).

In overview, each generation proceeds with females paying costs related to their reproductive strategy. Offspring inherit a randomly sampled allele from each parent at each locus. Alleles can then mutate and offspring express inbreeding depression in viability. Immigrants arrive in the population and density regulation limits population growth. We record the population pedigree and directly calculate the coefficient of kinship ($k$) between all potential mates in each generation (defined as the probability that two homologous alleles will be identical-by-descent), allowing individual pre-copulatory and post-copulatory inbreeding strategies to be enacted. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values have costs that are assumed to combine independently to increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding avoidance strategies because both negative and positive $M_{p}$ and $F_{p}$ values could potentially arise and affect the degree of inbreeding, representing inbreeding avoidance and inbreeding preference, respectively. Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with the phenotypic value of each trait.

@@

After costs are realised, each remaining female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate, and causes some stochastic variation around the expected mean $N_{males}$ of $P_{p}+1$.

@@

Most often, $N_{males}$ will be smaller than the total number of available males [@Duthie]. Each female then chooses her $N_{males}$ mates based on her pre-copulatory inbreeding avoidance phenotype ($M_{p}$), which equals her genotype ($M_{g}$). Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively, whereas $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

The probability that an offspring of female $i$ is sired by any one of $i$'s mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if female $i$ has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so a female $i$ samples from her mates $n$ times independently and with replacement with a probability of $q^{f}_{ij}$ for each male $j$ to determine the realised distribution of sires. Offspring produced have equal probability of being female or male.

@@

Offsprings' alleles mutate with independent probabilities $\mu=0.001$. When a mutation occurs, a mutation effect is sampled from a normal distribution with a mean of zero and a standard deviation of $\mu_{SD}$ and added to the original allele value [@Kimura1965; @Lande1976; @Bocedi2014; @Duthie]. The value of $\mu_{SD}$ is set to $\sqrt{1/20}$.

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, and not evolution where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a], or given outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are permitted to evolve, and could potentially arise due to mutation or drift.

@@

To address whether or not evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance can facilitate evolution of polyandry, we compare simulations in which polyandry and pre-copulatory and post-copulatory inbreeding avoidance can all evolve, with otherwise identical simulations in which post-copulatory inbreeding avoidance cannot evolve given four different costs of polyandry ($c_{P} = \{0, 0.0025, 0.005,  0.01\}$). To prevent post-copulatory inbreeding avoidance from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes cause random fertilisation with respect to kinship, so $F_{a}$ alleles have no phenotypic effect.

>>>>

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply 'inbreeding strategies' because the model did not preclude evolution of inbreeding preference), in a small focal population of females and males by tracking the dynamics of alleles underlying reproductive strategies expressed by females. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and allowing for effects of mutation, gene flow, drift, and selection.

All individuals have 10 physically unlinked diploid loci, and therefore 20 alleles, underlying each of three reproductive strategy traits (i.e., 30 diploid loci in total): tendency for polyandry ($P_{a}$, i.e., 'polyandry' alleles), pre-copulatory inbreeding strategy ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, i.e., 'fertilisation' alleles). All alleles can take the value of any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to determine genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and resulting phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy, respectively. Each individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. Each individual's phenotypic values for pre-copulatory and post-copulatory inbreeding strategy equal their respective genotypic values ($M_{p} = M_{g}$ and $F_{p} = F_{g}$), where negative and positive values cause inbreeding avoidance and preference, respectively (see details of mating and fertilisation strategies below). In contrast, phenotypic values for tendency for polyandry equal genotypic values ($P_{p} = P_{g}$) only if $P_{g} \geq 0$, while $P_{p} = 0$ if $P_{g} < 0$ because females cannot choose a negative number of mates. Polyandry is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998; @Duthie] that is influenced by continuous genetic variation but only expressed when $P_{g} > 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below).

In overview, each generation proceeds with females paying costs associated with their reproductive strategy traits, and expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus. Alleles can then mutate and offspring express inbreeding depression in viability. Immigrants arrive in the population and density regulation limits population growth. We record the population pedigree and directly calculate the coefficient of kinship ($k$) between all potential mates in each generation (defined as the probability that two homologous alleles will be identical-by-descent), allowing individual pre-copulatory and post-copulatory inbreeding strategies to be enacted. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values of the three reproductive strategy traits each have costs that are assumed to combine independently to increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding avoidance strategies because both negative and positive $M_{p}$ and $F_{p}$ values could potentially arise and affect the degree of inbreeding, representing inbreeding avoidance and inbreeding preference, respectively. In contrast, $P_{p}$ cannot be negative. Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with the phenotypic value of each trait.

@@

After costs are realised, each remaining female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate and generates each female's realised degree of polyandry with some stochastic variation around the expected mean $N_{males}$ of $P_{p}+1$.

@@

Most often, $N_{males}$ will be smaller than the total number of available males [@Duthie]. Each female then chooses her $N_{males}$ mates based on her pre-copulatory inbreeding avoidance phenotype ($M_{p}$). Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively, whereas $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

The probability that an offspring of female $i$ is sired by any one of $i$'s mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if female $i$ has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so a female $i$ samples from her mates $n$ times independently and with replacement with a probability of $q^{f}_{ij}$ for each male $j$ to determine the realised distribution of sires. Offspring have equal probability of being female or male.

@@

Offsprings' alleles mutate with independent probabilities $\mu=0.001$. When a mutation occurs, a mutation effect is sampled from a normal distribution with a mean of zero and a standard deviation of $\mu_{SD}$ and added to the original allele value [@Kimura1965; @Lande1976; @Bocedi2014; @Duthie]. The value of $\mu_{SD}$ is set to $\sqrt{1/20}$ because 20 independent alleles underlie each trait's genotypic value.

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, and not evolution where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a], or given outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

To address whether or not evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance can facilitate evolution of polyandry, we compare simulations in which polyandry and pre-copulatory and post-copulatory inbreeding avoidance can all evolve with otherwise identical simulations in which post-copulatory inbreeding avoidance cannot evolve. To achieve this, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes cause random fertilisation with respect to kinship, so $F_{a}$ alleles have no phenotypic effect. Simulations were repeated across four different costs of polyandry ($c_{P} = \{0, 0.0025, 0.005,  0.01\}$).
